### 2020-02-18 Writing/disparity_review.tex
reformated for tex compatible

RH: disparate disparity analyses.

\textbf{Disparities in the analysis of morphological disparity}

Thomas Guillerme\({}^{1,*, +}\), Natalie Cooper\({}^{2, +}\), Stephen L.
Brusatte\({}^{3}\), Katie E. Davis\({}^{4}\), Andrew L
Jackson\({}^{5}\), Sylvain Gerber\({}^{6}\), Anjali Goswami\({}^{2}\),
Kevin Healy\({}^{7}\), Melanie Hopkins\({}^{8}\), Marc EH
Jones\({}^{9}\), Graeme T. Lloyd\({}^{10}\), Joseph E.
O'Reilly\({}^{11}\), Abi Pate\({}^{11}\), Mark N Puttick\({}^{12}\),
Emiliy Rayfield\({}^{11}\), Erin E. Saupe\({}^{13}\), Emma
Sherratt\({}^{14}\), Graham Slater\({}^{15}\), Vera
Weisbecker\({}^{1}\), Gavin H Thomas\({}^{16}\) and Philip
Donoghue\({}^{11}\)\\
\({}^{1}\)\emph{School of Biological Sciences, University of Queensland,
St. Lucia, Queensland, Australia.}\\
\({}^{2}\)\emph{Department of Life Sciences, Natural History Museum,
London, Cromwell Road, London, SW7 5BD, UK.}

\emph{\textsuperscript{3}School of GeoSciences, University of Edinburgh,
Grant Institute, Edinburgh EH9 3FE, UK}

\textsuperscript{4} Department of Biology, University of York, YO10 5DD

\textsuperscript{5} Department of Zoology, School of Natural Sciences,
Trinity College.

\textsuperscript{6} Institut de Systématique, Évolution, Biodiversité
(ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne
Université, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005
Paris, France.

\textsuperscript{7} Ryan Institute, School of Natural Sciences, National
University of Ireland Galway, Ireland.

\textsuperscript{10} \emph{School of Earth and Environment, University
of Leeds, Leeds LS2 9JT, UK}\\
\textsuperscript{11}\emph{School of Earth Sciences, University of
Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK}

\textsuperscript{12} \emph{Milner Centre for Evolution, University of
Bath, BA2 7AY}

\textsuperscript{13} \emph{Department of Earth Sciences, University of
Oxford, S Parks Road, Oxford OX1 3AN, UK}

\textsuperscript{14} \emph{School of Biological Sciences, The University
of Adelaide, Adelaide, South Australia 5005, Australia}

+ These authors contributed equally to the manuscript.

*\textbf{Corresponding author:} \emph{guillert@tcd.ie}

\textbf{NOTE TO ALL THE CO-AUTHORS:}

\textbf{-The word limit for this is 5000 words so we can't add any large
sections.}

\textbf{-There is no reference limit per se but please don't add too
many. I will sort out the references later so don't correct the
formatting of them here.}

\textbf{-Add your affiliation in the author list above}. The
affiliations count in the word count so please 1) keep it short (max 1
per person!) and 2) try to use one that's already there. Also add any
initials I've missed in your names.

- If in doubt, check formatting here:
\href{https://royalsociety.org/journals/authors/author-guidelines/}{{https://royalsociety.org/journals/authors/author-guidelines/}}

Thanks!

@@

\includegraphics[width=6.5in,height=4.94444in]{media/image1.png}

@@

\includegraphics[width=5.63542in,height=3.875in]{media/image2.png}3.

@@

% Stopped p.13

@@

\includegraphics[width=6.5in,height=5.26389in]{media/image3.png}

\emph{Figure methods: Morphospaces: different mathematical
representations of a} \emph{morphospace. A trait matrix can be
transformed into a distance matrix (e.g. in Close et al., 2015) or an
ordinated matrix (e.g. in Brusatte et al.} \emph{2008). In each case,
all three matrices represent the morphospace for the observations at
hand. Visualisation: different ways to represent the morphospace in 2D.
Visualisations can use either trait plots (directly from the trait
matrix); ``flattening'' ordination of the data (e.g. using
multidimensional scaling} \emph{forcing all the variance to be contained
in two dimensions); or ordination axis plots (directly from the
ordinated matrix).}

\hypertarget{b-disparity-indices}{%
\subsection{(b) Measuring disparity}\label{c-disparity-indices}}

@@

When considering only one dimension, disparity metrics can be used to
reflect the spread of the distribution (e.g. the range, quantiles or
variance) or its central tendency (i.e., mean, median or mode). Among
these metrics, some will have more attractive properties than others,
such as sensitivity to outliers. Range and mean are highly sensitive,
whereas quantiles, variance and median are less so, making them more or
less appropriate for different questions. For example, if we want to
characterise the extent of morphospace occupied by a group (e.g. does
group A occupy as much space as group B?), metrics related to the spread
of the group in the morphospace are most appropriate (e.g. volume
\href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, distance
from the centroid
\href{https://paperpile.com/c/sTGYvp/vTHS+yyNa}{(Melanie J. Hopkins and
Gerber 2017; Finlay and Cooper 2015)}, variance and range
\href{https://paperpile.com/c/sTGYvp/tGyd}{(S. L. Brusatte et al.
2008)}). Conversely, if we wish to describe the ``position'' of a group
in a morphospace (e.g. does group A occupy the same space as group B?),
metrics related to the distance between the elements within a group and
a fixed point in the morphospace are most appropriate
(e.g.\href{https://paperpile.com/c/sTGYvp/u1KE}{(Guillerme et al.,
n.d.)}). Finally, if we aim to characterise the density of morphospace
occupation (e.g. is group A more closely packed than group B?) metrics
related to the pairwise distances between elements will be most
appropriate (e.g. nearest neighbour distance, pairwise distances, etc.
\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}).

In addition to considering what properties of disparity metrics should
capture, it is important to also consider the mathematical properties of
the metrics and their associated caveats
\href{https://paperpile.com/c/sTGYvp/nFf7+ROH8}{(Wills 2001; Ciampaglio,
Kemp, and McShea 2001)}. For example, measuring the full sum of the
variance in each dimension of the space does not require we add the
covariance between the axes in an ordinated space using a PCA. However,
this is not true of other mathematical spaces or when not all dimensions
or elements are considered, even in a PCA {[}Legendre \& Legendre
2012{]}.

Furthermore, multidimensional spaces also have some counter-intuitive
properties that need to be considered such as the ``curse of
dimensionality'' \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman
1966)}. In spaces with some axis of variance lower than 1, product-based
metrics used as proxies of volumes (e.g. product of ranges, hypervolume,
hypercube, etc.) can tend towards zero fairly quickly for spaces with
even a modest number of dimensions
\href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho
2000). Some other types of metrics are also extremely sensitive to
outliers and can be biased easily by sample size, for example range
\href{https://paperpile.com/c/sTGYvp/aSSL}{(}Butler et al.
2012\href{https://paperpile.com/c/sTGYvp/aSSL}{)} or convex hull based
metrics \href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al.
2012; Jackson et al. 2011)}.

@@

No matter which disparity metrics have been calculated, the research
question must be framed in an appropriate statistical context. The
multidimensional statistical toolkit for ecology and evolution has been
greatly expanded in recent years
\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer}
2018\href{https://paperpile.com/c/sTGYvp/ZnDd}{)}, but some of these
advances have yet to be implemented in disparity analyses. Instead,
hypothesis testing has mostly been confined to a small set of
well-established methods. One commonly used test is the non-parametric
permutation analysis of variance
\href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M. J. Anderson and
Walsh 2013; M. J. Anderson 2001)}, an analysis of variance (ANOVA) of
the pairwise distances between different groups. Although statistically
valid, this test is not always directly related to the hypothesis under
question. For example, PERMANOVA tests whether two groups share the same
variance/covariance in a ``distance-space''. This is not the same as
testing whether the two groups overlap in morphospace. Statistical tests
should be employed that are tailored to the question at hand, rather
than simply following common practices.

It is also important to consider which data should be subjected to a
statistical test. For example, in morphological disparity analysis,
especially for palaeobiological questions, data are often bootstrapped.
This has two advantages: (i) when the disparity metric is unidimensional
(e.g. the sum of variances), bootstrapping the data generates a
distribution of the metric that can be analysed using the vast
statistical toolkit available for comparing distributions; (ii) when
data are scarce, bootstrapping the data allows users to introduce
variance, rendering the test less sensitive to outliers. However,
bootstrapped data are pseudoreplicates and thus non-independent. This
violates the assumptions of most parametric statistical tests.
Furthermore, the number of bootstrap pseudoreplicates will inevitably
increase the Type I error rate. These factors are often ignored in
disparity analyses.

@@

As with all comparative datasets, the data used in disparity analyses
are not independent because close relatives will tend to have more
similar morphologies than more distant relatives
\href{https://paperpile.com/c/sTGYvp/WXik}{(Harvey and Pagel 1998)}.
Thus, for many disparity analyses, phylogenetic relationships should be
taken into account. However, it has been noted that some popular
phylogenetic correction methods (like pPCA) can be inappropriate if
incorrect assumptions are made about the data (Uyeda et al).
\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)}
provide a thorough review of multivariate phylogenetic comparative
methods, and so we do not consider them further here.

@@

No ancestral state estimation method is without drawbacks and above all
else are highly dependent on the data and method used. In general, using
ancestral state estimation can help with recovering patterns of changes
in disparity but should not be used simply to generate extra data points
to increase statistical power. In fact, these extra points are not
independent and can also have problematic side effects, especially when
testing for the influence of mass extinctions on disparity.

@@

Morphological disparity analyses are widely employed in evolutionary
palaeobiology, and they are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity
analyses. As with any multidimensional analysis, there are many
variables that have to be considered when deciding which data to use and
how to analyse it, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from
``blind'' application of established methodological pipelines without
consideration of the biological question being addressed. We advocate
the bespoke assembly of analytic approaches and, given the computational
efficiency of these methods, an experimental approach that explores the
impact of competing approaches, such as choice of distance measure,
ordination method and ancestral state estimation method on disparity
analysis results. Many of the methods employed in disparity analysis are
used more widely in other fields, including genomics and ecology, which
also encompass analyses of multidimensional datasets
\href{https://paperpile.com/c/sTGYvp/krNU+60H0+cV3v}{(Donohue et al.
2013; Canter et al. 2018; Saupe et al. 2015)}. Innovations in
morphological disparity analyses likely await discovery in their
respective literatures.

While studies of morphological disparity would benefit from advances in
multidimensional analysis in other fields, the concept of a morphospace
could reciprocally benefit other disciplines. For example, the
multidimensional analysis of Diaz
\href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, which
analysed patterns of form and function in plants, is essentially an
ecomorphospace; isotopic analyses of organisms
\href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al. 2011)} can be
represented as an isotope-space; ecosystem functioning in Donohue et al.
(2013) as an ecosystem-space, etc. These generalisations could also be
exported for any set of traits (e.g. acousto-spaces for acoustic traits
or glotto-spaces for linguistic traits). Cognate approaches have been
adopted recently in the analysis of single cell comparative
transcriptome data
\href{https://paperpile.com/c/sTGYvp/856K}{(Sebé-Pedrós et al. 2018)}
where interpretation of the resulting transcriptome-spaces would be
improved by heeding the concerns we highlight concerning morphospaces.

Although disparity analyses are now simple to implement in freely
available softwares
\href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{(e.g.,
Guillerme 2018;} Adams and Otárola-Castillo
2013\href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{;
Bouxin 2005; Harmon et al. 2008; Lloyd 2016; Navarro 2003; Dixon 2003)},
it is crucial to remember that they are multidimensional analyses and
multidimensional analyses are complex. We assert that future
morphological analyses will benefit by emphasising the methodological
decisions made, rather than simply using disparity analysis because
\emph{we can}.

@@

Donoho, D. L. (2000). High-dimensional data analysis: The curses and
blessings of dimensionality. \emph{AMS math challenges lecture},
\emph{1}(2000), 32.

References

\href{http://paperpile.com/b/sTGYvp/ZnDd}{Adams, Dean C., and Michael L.
Collyer. 2018. ``Multivariate Phylogenetic Comparative Methods:
Evaluations, Comparisons, and Recommendations.'' \emph{Systematic
Biology} 67 (1): 14--31.}

\href{http://paperpile.com/b/sTGYvp/J2G1}{Adams, Dean C., and Erik
Otárola-Castillo. 2013. ``Geomorph: Anrpackage for the Collection and
Analysis of Geometric Morphometric Shape Data.'' \emph{Methods in
Ecology and Evolution / British Ecological Society} 4 (4): 393--99.}

\href{http://paperpile.com/b/sTGYvp/SC6L}{Anderson, Marti J. 2001. ``A
New Method for Non-Parametric Multivariate Analysis of Variance.''
\emph{Austral Ecology} 26 (1): 32--46.}

\href{http://paperpile.com/b/sTGYvp/3hy2}{Anderson, Marti J., and Daniel
C. I. Walsh. 2013. ``PERMANOVA, ANOSIM, and the Mantel Test in the Face
of Heterogeneous Dispersions: What Null Hypothesis Are You Testing?''
\emph{Ecological Monographs} 83 (4): 557--74.}

\href{http://paperpile.com/b/sTGYvp/qjj9}{Anderson, Philip S. L., Matt
Friedman, Martin D. Brazeau, and Emily J. Rayfield. 2011. ``Initial
Radiation of Jaws Demonstrated Stability despite Faunal and
Environmental Change.'' \emph{Nature} 476 (7359): 206--9.}

\href{http://paperpile.com/b/sTGYvp/Qsl3}{Bellman, R. 1966. ``Dynamic
Programming.'' \emph{Science} 153 (3731): 34--37.}

\href{http://paperpile.com/b/sTGYvp/1SD2}{Bookstein, Fred L. 1997.
\emph{Morphometric Tools for Landmark Data: Geometry and Biology}.
Cambridge University Press.}

\href{http://paperpile.com/b/sTGYvp/sN5d}{---------. 2015. ``The
Relation between Geometric Morphometrics and Functional Morphology, as
Explored by Procrustes Interpretation of Individual Shape Measures
Pertinent to Function.'' \emph{Anatomical Record} 298 (1): 314--27.}

\href{http://paperpile.com/b/sTGYvp/o4w7}{---------. 2017a. ``A Newly
Noticed Formula Enforces Fundamental Limits on Geometric Morphometric
Analyses.'' \emph{Evolutionary Biology} 44 (4): 522--41.}

\href{http://paperpile.com/b/sTGYvp/xaUx}{---------. 2017b. ``A Method
of Factor Analysis for Shape Coordinates.'' \emph{American Journal of
Physical Anthropology} 164 (2): 221--45.}

\href{http://paperpile.com/b/sTGYvp/9JdS}{Bouxin, Guy. 2005. ``Ginkgo, a
Multivariate Analysis Package.'' \emph{Journal of Vegetation Science:
Official Organ of the International Association for Vegetation Science}
16 (3): 355--59.}

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage{natbib}

%Packages
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{pdflscape}

\usepackage{color,soul}

\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
\newcolumntype{d}[1]{D{.}{.}{#1}}

\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\begin{document}
%Running head
\begin{flushright}
Version dated: \today
\end{flushright}

\bigskip
\noindent RH: disparate disparity analyses
\bigskip
\medskip
\begin{center}
\noindent{\Large \bf Disparities in the analysis of morphological disparity}
\bigskip

Thomas Guillerme$^{1,16,*,+}$,
Natalie Cooper$^{2,+}$,
Stephen L. Brusatte$^{3}$,
Katie E. Davis$^{4}$,
Andrew L Jackson$^{5}$,
Sylvain Gerber$^{6}$,
Anjali Goswami$^{2}$,
Kevin Healy$^{7}$,
Melanie Hopkins$^{8}$,
Marc EH Jones$^{9}$,
Graeme T. Lloyd$^{10}$,
Joseph E. O'Reilly$^{11}$,
Abi Pate$^{11}$,
Mark N Puttick$^{12}$,
Emiliy Rayfield$^{11}$,
Erin E. Saupe$^{13}$,
Emma Sherratt$^{14}$,
Graham Slater$^{15}$,
Vera Weisbecker$^{1}$,
Gavin H Thomas$^{16}$
and Philip Donoghue$^{11}$\\

\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
$^2$IDepartment of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
$^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
$^{4}$Department of Biology, University of York, YO10 5DD, UK;
$^{5}$Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
$^{6}$Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
$^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
$^{8}$ @@@;
$^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
$^{11}$School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
$^{12}$Milner Centre for Evolution, University of Bath, BA2 7AY UK;
$^{13}$Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
$^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
$^{15}$ @@@;
$^{16}$ @@@;\\
$^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding author: guillert@tcd.ie\\}

\end{center}

@@

%\includegraphics[width=6.5in,height=4.94444in]{media/image1.png}

@@

%\includegraphics[width=5.63542in,height=3.875in]{media/image2.png}3.

@@

@@

%\includegraphics[width=6.5in,height=5.26389in]{media/image3.png}

\emph{Figure methods: Morphospaces: different mathematical representations of a} \emph{morphospace.
% SG: The morphospace is the original trait space within which lie the specimens. Then one performs an ordination to visualize the space. So I would distinguish the morphospace from the ordination space. See also Mitteroecker and Huttegger (2009)
A trait matrix can be transformed into a distance matrix (e.g. in Close et al., 2015) or an ordinated matrix (e.g. in Brusatte et al.} \emph{2008).
In each case, all three matrices represent the morphospace for the observations at hand.
Visualisation: different ways to represent the morphospace in 2D.
Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using multidimensional scaling} \emph{forcing all the variance to be contained in two dimensions)
% SG: No, the flattening will not force all the variance to be contained in 2D. First because NMDS is non metric and variance is therefore not preserved as such. Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation. Some information (variance) is thus necessarily lost.
; or ordination axis plots (directly from the ordinated matrix).}
% ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
% SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..). But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
% @@@ Change references

\hypertarget{b-disparity-indices}{%
% @@@ Change to indices throughout
\subsection{(b) Measuring disparity}\label{c-disparity-indices}}

@@

When considering only one dimension, disparity metrics can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these metrics, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), metrics related to the spread of the group in the morphospace are most appropriate (e.g. volume \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, distance from the centroid \href{https://paperpile.com/c/sTGYvp/vTHS+yyNa}{(Melanie J. Hopkins and Gerber 2017; Finlay and Cooper 2015)}, variance and range \href{https://paperpile.com/c/sTGYvp/tGyd}{(S. L. Brusatte et al. 2008)}).
Conversely, if we wish to describe the ``position''
% SG: Note that conceptually and technically,  disparity is about variation. So information on position is another aspect of morphospace occupation. but it is distinct from the meaning of  disparity as originally intended.
 of a group in a morphospace (e.g. does group A occupy the same space as group B?), metrics related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate (e.g.\href{https://paperpile.com/c/sTGYvp/u1KE}{(Guillerme et al., n.d.)}).
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) metrics related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}).
%ES: this para is probably most useful to refer to for the ordinate or not part

In addition to considering what properties of disparity metrics should capture, it is important to also consider the mathematical properties of the metrics and their associated caveats \href{https://paperpile.com/c/sTGYvp/nFf7+ROH8}{(Wills 2001; Ciampaglio, Kemp, and McShea 2001)}.
For example, measuring the full sum of the variance in each dimension of the space does not require we add the covariance between the axes in an ordinated space using a PCA.
% SG: Not very clear to me
However, this is not true of other mathematical spaces or when not all dimensions or elements are considered, even in a PCA {[}Legendre \& Legendre 2012{]}.

Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}.
In spaces with some axis of variance lower than 1, product-based metrics used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho 2000).
Some other types of metrics are also extremely sensitive to outliers and can be biased easily by sample size, for example range \href{https://paperpile.com/c/sTGYvp/aSSL}{(}Butler et al. 2012\href{https://paperpile.com/c/sTGYvp/aSSL}{)}
%SG: Or even better, twenty years earlier, Foote (1992), who orginally described the problem in the context of disparity and suggested a rarefaction approach.
or convex hull based metrics \href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al. 2012; Jackson et al. 2011)}.

@@

No matter which disparity metrics have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer} 2018\href{https://paperpile.com/c/sTGYvp/ZnDd}{)}
%AG: there are others that should be mentioned - mvmorph for example, which is especially important given that there are extremely different approaches to multivariate data in these different packages
, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance \href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M. J. Anderson and Walsh 2013; M. J. Anderson 2001)}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
Although statistically valid, this test is not always directly related to the hypothesis under question.
For example, PERMANOVA tests whether two groups share the same variance/covariance in a ``distance-space''.
% SG: What is meant by distance-sapce and in what way is it distinguished from the spaces discussed above?
This is not the same as testing whether the two groups overlap in morphospace.
%SG: You're sure of this?
Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.

It is also important to consider which data should be subjected to a statistical test.
% AG: It would be good to also note here that data should be comparable in terms of having a common reference frame - e.g. comparing disparities across elements with different numbers of landmarks or different landmark configurations that haven't been/can't be superimposed together is problematic but is popping up in the disparity literature
For example, in morphological disparity analysis, especially for palaeobiological questions, data are often bootstrapped.
This has two advantages: (i) when the disparity metric is unidimensional (e.g. the sum of variances), bootstrapping the data generates a distribution of the metric that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
However, bootstrapped data are pseudoreplicates and thus non-independent.
This violates the assumptions of most parametric statistical tests.
Furthermore, the number of bootstrap pseudoreplicates will inevitably increase the false positive rate (Type I error).
% @@@ check positive/negative type
These factors are often ignored in disparity analyses.
% VW: Surely there's a reference for this?
% MH: I might be wrong about this, but my impression is that CIs generated from bootstrapping are compared visually (like to assess significance in changes in disparity over time), but that statistical tests are not commonly applied.

@@

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \href{https://paperpile.com/c/sTGYvp/WXik}{(Harvey and Pagel 1998)}.
Thus, for many disparity analyses, phylogenetic relationships should be
taken into account.
% SG: Why?
However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate
% SG: In what way?
if incorrect assumptions are made about the data (Uyeda et al). \href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)}
% @@@ref: https://academic.oup.com/sysbio/article/64/4/677/1649888
% SG: Which assumption? This part (e) seems a bit vague to me
% AG: as long as you are using most or ideally all of the variance in a dataset, the rotation of pPCA should have little to no impact on disparity analyses. polly et al 2013 i think
provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
% AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly

@@

No ancestral state estimation method is without drawbacks
% @@@ Maybe check in Pennel’s “fossils are needed” paper for ref.
 and above all else are highly dependent on the data and method used.
In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points to increase statistical power.
In fact, these extra points are not independent and can also have problematic side effects, especially when testing for the influence of mass extinctions on disparity.

@@

Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse it, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate the bespoke
%VW: I think there could be just a line or two that takes makes "bespoke design" less daunting. There are loads of well-documented R packages available in morphological and ecological studies that can be relatively straightforwardly adapted.
assembly of analytic approaches and, given the computational efficiency of these methods, an experimental approach that explores the impact of competing approaches, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \href{https://paperpile.com/c/sTGYvp/krNU+60H0+cV3v}{(Donohue et al. 2013; Canter et al. 2018; Saupe et al. 2015)}. % @@@ Add Mammola
Innovations in morphological disparity analyses likely await discovery in their respective literatures.
% AG: as above, i think you need to add in evodevo work here, it's really fundamental to disparity

While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
For example, the multidimensional analysis of Diaz \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, which analysed patterns of form and function in plants, is essentially an ecomorphospace; isotopic analyses of organisms \href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al. 2011)} can be represented as an isotope-space; ecosystem functioning in Donohue et al. (2013) as an ecosystem-space, etc.
% AJ: this is a potential nice, although rather data hungry I believe, methods that I think could be worth dropping into the paper somewhere. Either here as a quick drop, or earlier up under overlap seems appropriate if you want to add it https://esajournals.onlinelibrary.wiley.com/doi/10.1890/14-0235.1
These generalisations could also be exported for any set of traits (e.g. acousto-spaces for acoustic traits or glotto-spaces for linguistic traits).
%SG: devising 
Cognate approaches have been adopted recently in the analysis of single cell comparative transcriptome data \href{https://paperpile.com/c/sTGYvp/856K}{(Sebé-Pedrós et al. 2018)} where interpretation of the resulting transcriptome-spaces would be improved by heeding the concerns we highlight concerning morphospaces.
% ES: Suggested by Erin: reword this sentence

Although disparity analyses are now simple to implement in freely available softwares \href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{(e.g., Guillerme 2018;} {; Bouxin 2005; Harmon et al. 2008; Lloyd 2016; Navarro 2003; Oksanen 2007)},
% @@@ add BAT
it is crucial to remember that they are multidimensional analyses and multidimensional analyses are complex.
We assert that future morphological analyses will benefit by emphasising the methodological decisions made, rather than simply using disparity analysis because \emph{we can}.

@@

\end{document}

% Donoho, D. L. (2000). High-dimensional data analysis: The curses and
% blessings of dimensionality. \emph{AMS math challenges lecture},
% \emph{1}(2000), 32.

% References

% \href{http://paperpile.com/b/sTGYvp/ZnDd}{Adams, Dean C., and Michael L.
% Collyer. 2018. ``Multivariate Phylogenetic Comparative Methods:
% Evaluations, Comparisons, and Recommendations.'' \emph{Systematic
% Biology} 67 (1): 14--31.}

% \href{http://paperpile.com/b/sTGYvp/J2G1}{Adams, Dean C., and Erik
% Otárola-Castillo. 2013. ``Geomorph: Anrpackage for the Collection and
% Analysis of Geometric Morphometric Shape Data.'' \emph{Methods in
% Ecology and Evolution / British Ecological Society} 4 (4): 393--99.}

% \href{http://paperpile.com/b/sTGYvp/SC6L}{Anderson, Marti J. 2001. ``A
% New Method for Non-Parametric Multivariate Analysis of Variance.''
% \emph{Austral Ecology} 26 (1): 32--46.}

% \href{http://paperpile.com/b/sTGYvp/3hy2}{Anderson, Marti J., and Daniel
% C. I. Walsh. 2013. ``PERMANOVA, ANOSIM, and the Mantel Test in the Face
% of Heterogeneous Dispersions: What Null Hypothesis Are You Testing?''
% \emph{Ecological Monographs} 83 (4): 557--74.}

% \href{http://paperpile.com/b/sTGYvp/qjj9}{Anderson, Philip S. L., Matt
% Friedman, Martin D. Brazeau, and Emily J. Rayfield. 2011. ``Initial
% Radiation of Jaws Demonstrated Stability despite Faunal and
% Environmental Change.'' \emph{Nature} 476 (7359): 206--9.}

% \href{http://paperpile.com/b/sTGYvp/Qsl3}{Bellman, R. 1966. ``Dynamic
% Programming.'' \emph{Science} 153 (3731): 34--37.}

% \href{http://paperpile.com/b/sTGYvp/1SD2}{Bookstein, Fred L. 1997.
% \emph{Morphometric Tools for Landmark Data: Geometry and Biology}.
% Cambridge University Press.}

% \href{http://paperpile.com/b/sTGYvp/sN5d}{---------. 2015. ``The
% Relation between Geometric Morphometrics and Functional Morphology, as
% Explored by Procrustes Interpretation of Individual Shape Measures
% Pertinent to Function.'' \emph{Anatomical Record} 298 (1): 314--27.}

% \href{http://paperpile.com/b/sTGYvp/o4w7}{---------. 2017a. ``A Newly
% Noticed Formula Enforces Fundamental Limits on Geometric Morphometric
% Analyses.'' \emph{Evolutionary Biology} 44 (4): 522--41.}

% \href{http://paperpile.com/b/sTGYvp/xaUx}{---------. 2017b. ``A Method
% of Factor Analysis for Shape Coordinates.'' \emph{American Journal of
% Physical Anthropology} 164 (2): 221--45.}
